#Practical Machine Learning - Course Project 
This project aims to demonstrate the predictive power of Machine Learning techniques, by processing the provided wearable-technology data set, and using accelerometer data to predict specific manners in which exercises were performed.  


###Data 
The provided data set can be found [here](http://groupware.les.inf.puc-rio.br/har) and consists of 19622 observations of 160 variables.  

```{r echo=FALSE,message=FALSE}
#download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv","pml-training.csv")
#download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv","pml-testing.csv")
pmlTraining <-  read.csv("pml-training.csv", sep = ",", header = TRUE)
pmlTesting <- read.csv("pml-testing.csv", sep=",", header = TRUE)
```

67 of these variables contain N/A values for more than 95% of the observations, so they will be excluded from further analysis.  
```{r}
#exclude columns with a large amount of NA values
naCounts <-sapply(pmlTraining, function(y) sum(length(which(is.na(y)))))
pmlTraining <- pmlTraining[,naCounts / nrow(pmlTraining) < 0.95]
```

An additional 42 variables are factor variables with very little variance, so they will be excluded as well.  We can also exclude the "x" and "timestamp" variables, as they are for identification purposes and do not provide any predictive power. 

We will keep the classe variable as it is the dependent variable to be predicted. 

```{r}
#exclude factor columns
isFactor <- sapply(pmlTraining, function(y) is.factor(y))
isFactor["classe"] = FALSE
pmlTraining <- pmlTraining[,!isFactor]

#exclude timestamps
pmlTraining <- pmlTraining[, !(colnames(pmlTraining) %in% c("x", "raw_timestamp_part_1", "raw_timestamp_part_2"))]
```

Note that our dataset now contains 19622 observations of 55 variables, and that all cases are complete: 
```{r}
dim(pmlTraining)
table(complete.cases(pmlTraining))
```

Below are several statements to generate correlation matrixes for each sensor location (belt, arm, forearm, dumbbell).  In the interest of brevity, the results or each correlation plot is not explicitly included -- they can be reproduced independently if desired.   

The correlation plots show that most data points are statiscally uncorrelated with similar data points, suggesting that they can each provide additional predictive power in our final model.  Several exceptions to this: 

- "roll\_belt" vs "total\_accel\_belt"
- "pitch\_belt" vs "accel\_belt\_x", "accel\_belt\_z", "magnet\_belt\_x"
- "gyros\_arm\_x" vs "gyros\_arm\_y"
- "magnet\_arm\_y" vs "accel\_arm\_x", "accel\_arm\_z", "magnet\_arm\_x", "magnet\_arm\_z"
- "magnet\_arm\_z" vs "accel\_arm\_x", "magnet\_arm\_x", "magnet\_arm\_y"
- "gyros\_dumbbell\_x" vs "gyros\_dumbbell\_z"

```{r, eval=FALSE}
install.packages("corrplot")
library(corrplot)
corrplot(cor(pmlTraining[,grep("belt", names(pmlTraining))]), method="pie")
corrplot(cor(pmlTraining[,grep("_arm_", names(pmlTraining))]), method="pie")
corrplot(cor(pmlTraining[,grep("forearm", names(pmlTraining))]), method="pie")
corrplot(cor(pmlTraining[,grep("dumbbell", names(pmlTraining))]), method="pie")
```

In light of this, we will also exclude pitch\_belt, magnet\_arm\_y, and magnet\_arm\_z variables. 

```{r echo=FALSE}
pmlTraining <- pmlTraining[, !(colnames(pmlTraining) %in% c("pitch_belt", "magnet_arm_y", "magnet_arm_z"))]
```


###Training and Predicting
We will reserve 30% of the training data set to use for validation purposes, so that we can get an idea of the efficacy of our algorithm before attempting to apply it to the testing data set.  

```{r}
# split the data 70 training /30 validating
library(caret)
set.seed(54321)
inTrain <- createDataPartition(pmlTraining$classe, p = .7, list = FALSE)
trainData <- pmlTraining[inTrain,]
validationData <- pmlTraining[-inTrain,]
```

Next, we can use the training data set to build a random forest predictive model.  We can then use this model to predict the 'classe' values in our validation data set.  

```{r}
library(randomForest)
#trainRF <- randomForest(classe~total_accel_belt+total_accel_arm+total_accel_dumbbell+total_accel_forearm, trainData)
trainRF <- randomForest(classe~., trainData)
validationPredRF <- predict(trainRF, validationData)
```


###Results 
```{r}
confusionMatrix(validationPredRF, validationData$classe)
```
